Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-09 paper2.qmd
section:theoretical; process:write,figures,code

<!-- 4. segway for next section -->
In aggregate, the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates. This overreliance, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, may also undermine the validity of these estimates [@Perron_et_al_2015, pp. 2]. Thus, addressing these issues through a more systematic and integrated approach could provide significant statistical and measurement benefits. For this reason, the study employs the structural approach to causal inference, as it fulfills these requirements.

@@

# Extending Thurstone's model {#sec-theoretical}

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example helps to clarify these concepts. The example will appear throughout the document to illustrate various aspects of the proposed extension. In this scenario, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they could design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges would then compare pairs of students' written texts (stimuli), producing a dichotomous outcome that reflects the relative quality of each text (trait). Based on this setup, researchers could reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Following current CJ practices, researchers would then rely on estimates from the BTL model, or its transformations, to approximate this estimand.

@@

<!-- 5. What the section will do? -->
As a result, this section addresses the identified limitations by extending Thurstone's general form (see @tbl-thurstone_cases) through a systematic and integrated approach based on the structural approach to causal inference. Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.

<!-- 6. What are the advantages? -->
In addition to addressing the identified limitations, the extended model offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current data analysis paradigm from passively accepting CJ's model assumptions to actively testing them with the specific data under analysis. For more details on these assumptions, refer to @sec-theory-issues.

@@

<!-- general development -->
<!-- recording: Sven 24.10.04; time: 00:21:00 - 00:34:10 -->
<!-- recording: Sven 24.10.04; time: 00:41:00 - 00:46:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:00:00 - 00:04:05 -->
<!-- recording: Sven and Tine 24.11.25; time: 01:00:00 - 01:08:40 -->
<!-- recording: Sven and Tine 24.12.17; time: 00:26:10 - 00:54:00 -->
<!-- recording: Sven and Tine 25.01.30 p1; time: 00:23:25 - 00:48:00 -->

<!-- 1. The SCMs -->
Before incorporating the theoretical principles into the *conceptual-population model*, the study defines SCMs, which are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other endogenous and exogenous variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

@@

<!-- ![](/images/png/CJ_TM_01.png){#fig-cj01_dag fig-align="center" fig-pos="H" width=23%} -->

<!-- ![](/images/png/CJ_TM_02.png){#fig-cj02_dag fig-align="center" fig-pos="H" width=32%} -->

<!-- ![](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=45%} -->

::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{abhijk} & := f_{O}(D_{abhijk}) \\
  D_{abhijk} & := f_{D}(PT_{ahjk}, PT_{bijk}) \\
  PT_{ahjk} & := f_{PT}(T_{ah}, B_{jk}) \\
  PT_{bijk} & := f_{PT}(T_{bi}, B_{jk})
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=45%}

@@

::: {#fig-cj05 layout-nrow=2}

::: {#fig-cj05_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ})
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_05.png){#fig-cj05_dag fig-align="center" fig-pos="H" width=44%}

@@

<!-- ![](/images/png/CJ_TM_06.png){#fig-cj06_dag fig-align="center" fig-pos="H" width=59%} -->

<!-- ![](/images/png/CJ_TM_07.png){#fig-cj07_dag fig-align="center" fig-pos="H" width=68%} -->

<!-- ![](/images/png/CJ_TM_08.png){#fig-cj08_dag fig-align="center" fig-pos="H" width=68%} -->

<!-- ![](/images/png/CJ_TM_09.png){#fig-cj09_dag fig-align="center" fig-pos="H" width=68%} -->

::: {#fig-cj10 layout-nrow=2}

::: {#fig-cj10_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_10.png){#fig-cj10_dag fig-align="center" fig-pos="H" width=68%}

@@

<!-- ![](/images/png/CJ_TM_11.png){#fig-cj11_dag fig-align="center" fig-pos="H" width=80%} -->

::: {#fig-cj12 layout-nrow=2}

::: {#fig-cj12_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{S}_{PRQ} & := f_{O}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J})
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_12.png){#fig-cj12_dag fig-align="center" fig-pos="H" width=80%}

@@

<!-- ![](/images/png/CJ_TM_13.png){#fig-cj13_dag fig-align="center" fig-pos="H" width=90%} -->

::: {#fig-cj14 layout-nrow=2}

::: {#fig-cj14_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_14.png){#fig-cj14_dag fig-align="center" fig-pos="H" width=90%}

::: {#fig-cj15 layout-nrow=2}

::: {#fig-cj15_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{C}_{PRQ} & := f_{C}(O^{S}_{PRQ}, C) \\
  O^{S}_{PRQ} & := f_{S}(O_{PRQ}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O_{PRQ} & := f_{O}(D_{PRQ}) \\
  D_{PRQ} & := f_{D}(PT_{PRQ}) \\
  PT_{PRQ} & := f_{PT}(T_{AI}, B_{KJ}) \\
  T_{AI} & := f_{T}(T_{I}, X_{A}, e_{A}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{KJ} & := f_{B}(B_{J}, Z_{K}, e_{K}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
$$

![DAG](/images/png/CJ_TM_15.png){#fig-cj15_dag fig-align="center" fig-pos="H" width=90%}

@@

**Funding:** The project was founded through the Research Fund of the University of Antwerp (BOF).

**Financial interests:** The authors have no relevant financial interest to disclose.

**Non-financial interests:** The authors have no relevant non-financial interest to disclose.

**Ethics approval:** The University of Antwerp Research Ethics Committee has confirmed that no ethical approval is required.

@@

**Consent for publication:** All authors have read and agreed to the published version of the manuscript.

**Availability of data and materials:** No data was utilized in this study.

**Code availability:** All the code utilized in this research is available in the digital document located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/).

**AI-assisted technologies in the writing process:** The authors utilized a range of AI-based language tools throughout the preparation of this work. They occasionally employed the tools to refine phrasing and optimize wording, ensuring appropriate language use and enhancing the manuscript's clarity and coherence. The authors take full responsibility for the final content of the publication.

@@

@@

## Empirical research and randomized experiments {#sec-appendix-A .appendix appendix-style=plain}

@@

## Identification under causal inference {#sec-appendix-B .appendix appendix-style=plain}

<!-- ### The general idea {#sec-appendix-B1 .appendix appendix-style=plain} -->

@@

<!-- 2. What are counterfactuals? -->
*Counterfactuals* are hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals. For instance, in the scenario described in @sec-appendix-A, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes, as in @eq-ICE.

@@

For instance, if researchers cannot conduct the randomized experiments described in @sec-appendix-A and must instead rely on observational data, they can still identify the ACE as long as an observed variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion. Under these circumstances, researchers first identify the *conditional average causal effect* (CACE, @eq-CACE)

@@

### Diving into the specifics {#sec-appendix-B2 .appendix appendix-style=plain}

@@

SCMs and DAGs support identification analysis through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in a non-parametric and fully interactive way. This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp. 35]. @sec-appendix-B21 and @sec-appendix-B22 elaborate on the first advantage, while @sec-appendix-B22 and @sec-appendix-B23 do so for the second.

@@

#### The five fundamental block for SCMs and DAGs {#sec-appendix-B21 .appendix appendix-style=plain}

@@

<!-- 5. Continuing with the motivating example -->
Researchers can then use these building blocks to represent the scenario outlined in @sec-appendix-B. SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example. In this context, the variable $X$ (socio-economic status of the school) is thought to be a confounder in the relationship between the teaching method $T$ and the outcome $Y$. The figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. They also highlight unconnected node pairs, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Additional, the figures show one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.

@@

Plausible causal structure the scenario outlined in @sec-appendix-B.
:::

#### The probabilistic implications of these blocks {#sec-appendix-B22 .appendix appendix-style=plain}

@@

<!-- 4. can you illustrate this concept? -->
The five fundamental building blocks described in @sec-appendix-B21 clearly illustrate which (in)dependencies can SMCs and DAGs predict. For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] enables researchers to express the joint probability distribution of the observed variables as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, supporting the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$) [@Neal_2020, pp. 24]. Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \cond X_{1}) P(X_{1})$. Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendix-B21 and the CPDs.

@@

<!-- 5. Continuing with the motivating example -->
Using these additional probabilistic insights, researchers can revisit the scenario in @sec-appendix-B. In this context, applying the BNF to the SCM [-@fig-scm_example2] structure, enables the representation of the joint probability distribution of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$. From this expression, researchers can infer that the outcome $Y$ is unconditionally dependent on the teaching method $T$ $(Y \ndsep T)$. This dependency arises from two key structures: a direct causal path from the teaching method $T$ to the outcome $Y$, represented by the two-connected-nodes structure $T \rightarrow Y$ (black path in DAG [-@fig-mdag_example2]), and a confounding non-causal path from the teaching method $T$ to the outcome $Y$ through the socio-economic status of the school $X$, represented by the fork structure $T \leftarrow X \rightarrow Y$ (gray path in DAG [-@fig-mdag_example2]).

@@

Plausible causal structure the scenario outlined in @sec-appendix-B.
:::

#### From probability to causality {#sec-appendix-B23 .appendix appendix-style=plain}

@@

<!-- 4. revisiting the motivating example -->
With these new insights, researchers revisiting the scenario in @sec-appendix-B22 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] is correct. This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure. Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \cond X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$. Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example2] is equivalent to the manipulated DAG [-@fig-mdag_example3], because $X$ satisfies the backdoor criterion.

@@

Plausible causal structure the scenario outlined in @sec-appendix-B22.
:::

Researchers can then apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$. They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE. This process is expressed in @eq-mACE2 (see @sec-appendix-B).

@@

#### The estimation process {#sec-appendix-C .appendix appendix-style=plain}

@@

<!-- 3. How do DAGs can help? -->
Nevertheless, researchers can manage the complexity of the likelihood by assuming specific local (in)dependencies among variables. SCMs and DAGs provide a formal framework to represent these assumptions, as detailed in @sec-appendix-B22. These assumptions improve model tractability and simplify the estimation process by enabling the derivation of the BNF of the likelihood (@eq-like_BNF). With this simplified structure, any probabilistic programming language can model the system and compute the parameter's posterior distribution using @eq-bayes_rule.

>>>>

<!-- 4. segway for next section -->
In aggregate, the excessive reliance on Thurstone's Case V assumptions in the statistical analysis of CJ data can compromise the reliability of CJ's trait estimates. This overreliance, coupled with the apparent disconnect between CJ's approach to trait measurement and hypothesis testing, may also undermine the validity of these estimates [@Perron_et_al_2015, pp. 2]. Thus, addressing these issues through a more systematic and integrated approach could provide significant statistical and measurement benefits. For this reason, the study employs the structural approach to causal inference.

@@

# Extending Thurstone's general form {#sec-theoretical}

@@

<!-- 3. Example of a CJ estimand and estimator, and its limitations -->
A motivating example helps to clarify these concepts. This example will appear throughout the document to illustrate various aspects of the proposed extension. In this scenario, researchers aim to determine: "To what extent do different teaching methods influence students' ability to produce high-quality written texts?" To investigate this, they design a CJ experiment by randomly assigning students (individuals) to two groups, each receiving a different teaching method. Judges then compare pairs of students’ written texts (stimuli), producing a dichotomous outcome that reflects the relative quality of each text (trait). Based on this setup, researchers could reformulate the research question as the estimand: "*On average*, is there a difference in the ability to produce high-quality written texts between the two groups of students?". Following current CJ practices, researchers would then rely on estimates from the BTL model, or its transformations, to approximate this estimand.

@@

<!-- 5. What the section will do? -->
Thus, this section addresses the identified limitations by extending Thurstone’s general form (see @tbl-thurstone_cases) through a systematic and integrated approach based on the structural approach to causal inference. Specifically, it leverages the capabilities of this approach to formalize the combination of the core theoretical principles outlined in @sec-thurstone_theory with key assessment design features relevant to CJ experiments, such as the selection of judges, stimuli, and comparisons.

<!-- 6. What are the advantages? -->
In addition to overcoming the identified limitations, the extended model offers two key advantages. First, it clarifies the interactions among all actors and processes involved in CJ experiments. Second, it shifts the current comparative data analysis paradigm from passively accepting the model assumptions to actively testing if those assumptions fit the specific data under analysis. For more details on these assumptions, refer to @sec-theory-issues.

@@

<!-- general development -->
<!-- recording: Sven 24.10.04; time: 00:21:00 - 00:34:10 -->
<!-- recording: Sven 24.10.04; time: 00:41:00 - 00:46:00 -->
<!-- recording: Sven and Tine 24.10.18; time: 00:00:00 - 00:04:05 -->
<!-- recording: Sven and Tine 24.11.25; time: 01:00:00 - 01:08:40 -->
<!-- recording: Sven and Tine 24.12.17; time: 00:26:10 - 00:54:00 -->
<!-- recording: Sven and Tine 25.01.30 p1; time: 00:23:25 - 00:48:00 -->
<!-- recording: Sven and Tine 25.02.07; time: 00:00:00 - 00:00:00 -->

<!-- 1. The SCMs -->
Before incorporating the theoretical principles into the *conceptual-population model*, the study defines SCMs. SCMs are formal mathematical models characterized by a set of *endogenous* variables $V$, a set of *exogenous* variables $E$, and a set of functions $F$ [@Pearl_2009; @Cinelli_et_al_2020]. Endogenous variables are those whose causal mechanisms a researcher chooses to model [@Neal_2020]. In contrast, exogenous variables represent *errors* or *disturbances* arising from omitted factors that the investigator chooses not to model explicitly [@Pearl_2009, pp. 27,68]. Lastly, the functions, referred to as *structural equations*, express the endogenous variables as non-parametric functions of other endogenous and exogenous variables. These functions use the symbol '$:=$' to denote the asymmetrical causal dependence between variables and the symbol '$\dsep$' to represent *d-separation*, a concept akin to (conditional) independence.

@@

::: {#fig-cj03 layout-nrow=2}

::: {#fig-cj03_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{abhijk} & := f_{O}(D_{abhijk}) \\
  D_{abhijk} & := f_{D}(T_{ah}, T_{bi}, B_{jk})
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_03.png){#fig-cj03_dag fig-align="center" fig-pos="H" width=35%}

@@

::: {#fig-cj04_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK})
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_04.png){#fig-cj04_dag fig-align="center" fig-pos="H" width=44%}

@@

::: {#fig-cj09 layout-nrow=2}

::: {#fig-cj09_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_09.png){#fig-cj09_dag fig-align="center" fig-pos="H" width=70%}

@@

::: {#fig-cj11 layout-nrow=2}

::: {#fig-cj11_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_11.png){#fig-cj11_dag fig-align="center" fig-pos="H" width=84%}

@@

::: {#fig-cj13 layout-nrow=2}

::: {#fig-cj13_scm fig-align="center" fig-pos="H"}
$$
\begin{aligned}
  O_{RPUV} & := f_{C}(O^{sc}_{RPUV}, C) \\
  O^{sc}_{RPUV} & := f_{O}(O^{cp}_{RPUV}, S_{I}, S_{A}, S_{K}, S_{J}) \\
  O^{cp}_{RPUV} & := f_{O}(D_{RPUV}) \\
  D_{RPUV} & := f_{D}(T_{IA}, B_{JK}) \\
  T_{IA} & := f_{T}(T_{I}, X_{A}, e_{IA}) \\
  T_{I} & := f_{T}(X_{I}, e_{I}) \\
  B_{JK} & := f_{B}(B_{J}, Z_{K}, e_{JK}) \\
  B_{J} & := f_{B}(Z_{J}, e_{J}) 
\end{aligned}
$$

@@

![DAG](/images/png/CJ_TM_13.png){#fig-cj13_dag fig-align="center" fig-pos="H" width=100%}

@@

**Funding:** The Research Fund (BOF) of the University of Antwerp funded this project.

**Financial interests:** The authors declare no relevant financial interests.

**Non-financial interests:** The authors declare no relevant non-financial interests.

**Ethics approval:** The University of Antwerp Research Ethics Committee confirmed that this study does not require ethical approval.

@@

**Consent for publication:** All authors have read and approved the final version of the manuscript for publication.

**Data availability:** This study did not use any data.

**Materials and code availability:** The `CODE LINK` section at the top of the digital document located at: [https://jriveraespejo.github.io/paper2_manuscript/](https://jriveraespejo.github.io/paper2_manuscript/) provides access to all materials and code.

**AI-assisted technologies in the writing process:** The authors used various AI-based language tools to refine phrasing, optimize wording, and enhance clarity and coherence throughout the manuscript. They take full responsibility for the final content of the publication.

@@

## Statistics and Causal inference {#sec-appendixB .appendix appendix-style=plain}

@@

### Empirical research and randomized experiments {#sec-appendixB1 .appendix appendix-style=plain}

@@

### Identification under causal inference {#sec-appendixB2 .appendix appendix-style=plain}

@@

<!-- 2. What are counterfactuals? -->
*Counterfactuals* are hypothetical scenarios that are *contrary to fact*, where alternative outcomes resulting from a given cause are neither observed nor observable [@Neal_2020; @Counterfactual_2024]. The structural approach to causal inference [@Pearl_2009; @Pearl_et_al_2016] provides a formal framework for defining counterfactuals. For instance, in the scenario described in @sec-appendixB1, the approach begins by defining the *individual causal effect* (ICE) as the difference between each student's potential outcomes, as in @eq-ICE.

@@

For instance, if researchers cannot conduct the randomized experiments described in @sec-appendixB1 and must instead rely on observational data, they can still identify the ACE as long as an observed variable $X$, such as the socio-economic status of the school, satisfies the backdoor criterion. Under these circumstances, researchers first identify the *conditional average causal effect* (CACE, @eq-CACE)

@@

### Diving into the specifics {#sec-appendixB3 .appendix appendix-style=plain}

@@

SCMs and DAGs support identification analysis through two key advantages. First, regardless of complexity, they can represent various causal structures using only five fundamental building blocks [@Neal_2020; @McElreath_2020]. This feature allows researchers to decompose complex structures into manageable components, facilitating their analysis [@McElreath_2020]. Second, they depict causal relationships in a non-parametric and fully interactive way. This flexibility enables feasible ACE identification strategies without defining the variables' data types, the functional form between them, or their parameters [@Pearl_et_al_2016, pp. 35].

Thus, @sec-appendixB31 and @sec-appendixB32 elaborate on the first advantage, while @sec-appendixB32 and @sec-appendixB33 do so for the second. Finally, @sec-appendixB34 explains how researchers use SCMs and DAGs alongside Bayesian inference methods in the estimation process.

@@

#### The five fundamental block for SCMs and DAGs {#sec-appendixB31 .appendix appendix-style=plain}

@@

<!-- 5. Continuing with the motivating example -->
Researchers can then use these building blocks to represent the scenario outlined in @sec-appendixB2. SCM [-@fig-scm_example1] and DAG [-@fig-mdag_example1] depict the plausible causal structure for this example. In this context, the variable $X$ (socio-economic status of the school) is thought to be a confounder in the relationship between the teaching method $T$ and the outcome $Y$. The figures display multiple descendant relationships such as $X \rightarrow T$, $X \rightarrow Y$, and $T \rightarrow Y$. They also highlight unconnected node pairs, evident from the relationships $e_{T} \dsep e_{X}$, $e_{T} \dsep e_{Y}$, and $e_{X} \dsep e_{Y}$. Additional, the figures show one fork, $X \rightarrow \{T, Y\}$, and two colliders: $\{X, e_{T}\} \rightarrow T$ and $\{X, T, e_{Y}\} \rightarrow Y$.

@@

Plausible causal structure the scenario outlined in @sec-appendixB2.
:::

#### The probabilistic implications of these blocks {#sec-appendixB32 .appendix appendix-style=plain}

@@

<!-- 4. can you illustrate this concept? -->
The five fundamental building blocks described in @sec-appendixB31 clearly illustrate which (in)dependencies can SMCs and DAGs predict. For instance, applying the BNF to the causal structure shown in the SCM [-@fig-scm_bb1] and DAG [-@fig-mdag_bb1] enables researchers to express the joint probability distribution of the observed variables as $P(X_{1}, X_{3}) = P(X_{1}) P(X_{3})$, supporting the theoretical assumption that the observed variables $X_{1}$ and $X_{3}$ are unconditionally independent ($X_{1} \dsep X_{3}$) [@Neal_2020, pp. 24]. Conversely, when $X_{3}$ is unconditionally dependent on $X_{1}$ ($X_{1} \ndsep X_{3}$), as depicted in the SCM [-@fig-scm_bb2] and DAG [-@fig-mdag_bb2], the BNF express their joint probability distribution as $P(X_{1}, X_{3}) = P(X_{3} \cond X_{1}) P(X_{1})$. Notably, these descriptions demonstrate the clear correspondence between the structural equations illustrated in @sec-appendixB31 and the CPDs.

@@

<!-- 5. Continuing with the motivating example -->
Using these additional probabilistic insights, researchers can revisit the scenario in @sec-appendixB2. In this context, applying the BNF to the SCM [-@fig-scm_example2] structure, enables the representation of the joint probability distribution of the observed variables as $P(Y, T, X) =$ $P(Y \cond T, X) P(T \cond X) P(X)$. From this expression, researchers can infer that the outcome $Y$ is unconditionally dependent on the teaching method $T$ $(Y \ndsep T)$. This dependency arises from two key structures: a direct causal path from the teaching method $T$ to the outcome $Y$, represented by the two-connected-nodes structure $T \rightarrow Y$ (black path in DAG [-@fig-mdag_example2]), and a confounding non-causal path from the teaching method $T$ to the outcome $Y$ through the socio-economic status of the school $X$, represented by the fork structure $T \leftarrow X \rightarrow Y$ (gray path in DAG [-@fig-mdag_example2]).

@@

Plausible causal structure the scenario outlined in @sec-appendixB2.
:::

#### From probability to causality {#sec-appendixB33 .appendix appendix-style=plain}

@@

<!-- 4. revisiting the motivating example -->
With these new insights, researchers revisiting the scenario in @sec-appendixB32 can infer that the socio-economic status of the school, $X$, satisfies the backdoor criterion, assuming the causal structure depicted by the SCM [-@fig-scm_example2] and DAG [-@fig-mdag_example2] is correct. This means that $X$ serves as a sufficient adjustment set, as it effectively blocks all confounding non-causal paths introduced by the fork structure. Nevertheless, since $Y$ remains dependent on $T$ even after conditioning $(Y \ndsep T \cond X)$, this dependency can only be attributed to the direct causal effect $T \rightarrow Y$. Notably, for the purpose of identification, the conditioned DAG [-@fig-mdag_example2] is equivalent to the manipulated DAG [-@fig-mdag_example3], because $X$ satisfies the backdoor criterion.

@@

Plausible causal structure the scenario outlined in @sec-appendixB32.
:::

Researchers can then apply the *backdoor adjustment* to identify the ACE of $T$ on $Y$. They achieve this by first identifying the CACE of $T$ on $Y$ by conditioning on $X$, and then marginalizing this effect over $X$ to obtain the ACE. This process is expressed in @eq-mACE2 (see @sec-appendixB2).

@@

#### The estimation process {#sec-appendixB34 .appendix appendix-style=plain}

@@

<!-- 3. How do DAGs can help? -->
Nevertheless, researchers can manage the complexity of the likelihood by assuming specific local (in)dependencies among variables. SCMs and DAGs provide a formal framework to represent these assumptions, as detailed in @sec-appendixB32. These assumptions improve model tractability and simplify the estimation process by enabling the derivation of the BNF of the likelihood (@eq-like_BNF). With this simplified structure, any probabilistic programming language can model the system and compute the parameter's posterior distribution using @eq-bayes_rule.
